## Exercise 11

From Five sires (bulls used for breeding are often referred to as "sires") have we recoded the annual milk production from 8 daughters which are randomly picked out from all daughter of those sires. The data: `Sires.Rdata` (fronter)

```{r}
load("_data/Sires.Rdata")
Sires$sire <- as.factor(Sires$sire)
```

a) State a model with `milk` production as response and `sire` as factor. How would you interpret all the parameters in this model?

<div class = 'ans'>
Let $Y_{ij}$ be milk produced by daughter $j$ of sires $i$.

Model: $$Y_{ij} = \mu + \tau_i + \epsilon_{ij}$$

Here, $\epsilon_{ij} \sim \text{NID}(0, \sigma^2)$ and $i = 1, 2, \ldots 5$ and $j = 1, 2, \ldots 8$

Here, we also have one restriction,

$$\sum_{i = 1}^5\tau_i = 0$$

**Interpretaion of Parameters:**

|            |   |                                                                            |
|-----------:|:-:|----------------------------------------------------------------------------|
|      $\mu$ | : | Average milk production for all daughters under the 5 sires                |
|   $\tau_i$ | : | The average profit in annually milk production of using only sire $i$      |
| $\sigma^2$ | : | The variance in milk production for all daughter under one particular sire |

</div>


b) Test if there is effect of `sire` with respect to milk production. (R-commander: _Statistics > Fit model > Linear model_. Note that `sire` must be factor variable)

<div class = 'ans'>
The hypothesis is,

$$
\begin{aligned}
H_0 &: \tau_1 = \tau_2 = \tau_3 = 0 \\
H_1 &: \tau_i \ne 0 \text{ for at least one }i
\end{aligned}
$$ 

ANOVA for this model is,

```{r}
mdl <- lm(milk ~ sire, data = Sires)
aovsumry <- unclass(summary(aov(mdl)))[[1]]
sumry <- knitr::kable(aovsumry)
gsub('NA', '  ', sumry)
```

Here, sire effect is significant at 5% (p-value: `r round(aovsumry[1, "Pr(>F)"], 4)`)
</div>


c) Estimate all parameters in the model (R-commander: _Model > summarize model_)

<div class = 'ans'>

The summary of above fitted model is,

```{r}
(mdlSumry <- summary(mdl))
```

Here, the estimates of unknown parameters are,

```{r, results='asis'}
estdf <- data.frame(estimates = c(coef(mdl), sigma = mdlSumry$sigma))
rownames(estdf) <- c("$\\hat{\\mu}$", "$\\hat{\\tau}_1$", "$\\hat{\\tau}_2$", "$\\hat{\\tau}_3$","$\\hat{\\tau}_4$", "$\\hat{\\sigma}$")
knitr::kable(estdf)
```

</div>


d) (Use the general fact that in an one way ANOVA model with $a$ levels on the factor and $n$ replicates ${}^\text{SSE}\!/_{\sigma^2}$ is chi square distributed with $na – a$ degrees of freedom). Find a 95 % confidence interval for the error variance (the unexplained variance). Explain the interval to farmer. (use the general fact that in an one way ANOVA model with a levels on the factor and $n$ replicates ${}^\text{SSE}\!/_{\sigma^2}$ is chi square distributed with $na – a$ degrees of freedom)

<div class = 'ans'>
95% confidence interval for $\sigma^2$ is given by,


```{r}
sse <- round(aovsumry[['Sum Sq']][2], 1)
sigma.ci <- c(sse / qchisq(0.975, 35), sse / qchisq(0.025, 35))
```

$$
\begin{aligned}
\left(
\frac{\text{SSE}}{\chi^2_{35, 0.975}}, \frac{\text{SSE}}{\chi^2_{35, 0.025}}
\right) &=
\left(
\frac{`r sse`}{`r qchisq(0.975, 35)`}, \frac{`r sse`}{`r qchisq(0.025, 35)`}
\right) \\ &= 
\left(`r sigma.ci`\right)
\end{aligned}
$$ 

Therefore, 95% confidence limit for $\sigma$ is (`r round(sqrt(sigma.ci))`) kg of milk. This value revers that with 95% of confidence level, the true variation in annual milk production for all daughter under one particular sire lie between (`r round(sqrt(sigma.ci))`) kg.
</div>
