## Exercise 12

```{r}
dta1 <- `rownames<-`(matrix(c(69, 79, 89, 71, 81, 91), 3, 2), c("Diet1", "Diet2", "Diet3"))
dta2 <- `rownames<-`(matrix(c(60, 70, 80, 80, 90, 100), 3, 2), c("Diet1", "Diet2", "Diet3"))
dta3 <- `rownames<-`(matrix(c(70, 71, 72, 90, 91, 92), 3, 2), c("Diet1", "Diet2", "Diet3"))
dta4 <- `rownames<-`(matrix(c(80, 82, 83, 82, 84, 85), 3, 2), c("Diet1", "Diet2", "Diet3"))
dta <- list(dta1, dta2, dta3, dta4)
```

Tree diets for pigs were test out, in an experiment with 2 replicates. The Response is carcass weight (slaktevekt in Norwegian). Let’s assume 4students did the experiment and that they really obtained different results:

### Experiment 1

```{r}
knitr::kable(dta1)
```

### Experiment 2

```{r}
knitr::kable(dta2)
```

### Experiment 3

```{r}
knitr::kable(dta3)
```

### Experiment 4

```{r}
knitr::kable(dta4)
```

### Without computing anything:

<ul>
<li>Describe all 4 experiments, have focus on large explained or unexplained variation.
<div class = 'ans'>
**Experiment 1:** Large difference between diets, small difference inside diets 

**Experiment 2:** Large difference between diets, large difference inside diets 

**Experiment 3:** Small difference between diets, large difference inside diets 

**Experiment 4:** Small difference between diets, small difference inside diets
</div>
</li>
<li>In which experiment do you think you will have a large, small or moderate F value?
<div class = 'ans' id = "pfvalue">
**Experiment 1:** Large F, small P-value

**Experiment 2:** Moderate F

**Experiment 3:** Small F, large P-value 

**Experiment 4:** Moderate F
</div>
</li>
<li>In which experiment do you think you will have a large, small or moderate P value?
<div class = 'ans'>
See [previous answer](#pfvalue)
</div>
</li>
<li>In which experiment do you think you will prove a significant effect of diet?
<div class = 'ans'>
**Experiment 1** will prove a significant effect of diet due to large difference between diets and small difference inside diets.

**Experiment 2** could results with significant effect of diet but may get influenced by the noise within diets.

**Experiment 3** and **Experiment 4** has very small difference between diets however Experiment 4 has small variation within diets which can help us to see the difference between diets.
</div>
</li>
<li>Check your answers by R-commander.
<div class = 'ans'>
The ANOVA table for four experiments from R-commander is,

```{r}
dta.stack <- lapply(dta, function(x) {
  `names<-`(stack(as.data.frame(t(x))), c('weight', 'diet'))
})
mdls <- lapply(dta.stack, function(x) lm(weight ~ diet, data = x))
aovTbls <- lapply(mdls, function(x) {
  out <- knitr::kable(unclass(summary(aov(x)))[[1]])
  gsub('NA', '  ', out)
})
```

### Experiment 1:

```{r}
aovTbls[[1]]
```

### Experiment 2:

```{r}
aovTbls[[2]]
```

### Experiment 3:

```{r}
aovTbls[[3]]
```

### Experiment 4:

```{r}
aovTbls[[4]]
```



</div>
</li>
</ul>