## Exercise 4
Assume the following result (on scale 0-100) based on 8 randomly selected students from an exam in a basic course in statistics.

```{r}
lns <- "
Females 80 85 73 69
Males 90 60 76 67
"
con <- textConnection(lns)
df <- read.table(con, sep = " ")
close(con)
df <- `colnames<-`(`rownames<-`(df[, -1], df[, 1]), paste0('S', 1:4))
knitr::kable(df)
```

a) Execute a test (choose the level of significance yourself) to investigate if there is larger variation in statistical ability among males compared to females.

```{r, results='asis'}
sumry <- sapply(list(N = length, Mean = mean, Variance = var), function(fn){
  apply(df, 1, fn)
})
knitr::kable(sumry, format = 'html')
```

<div class = 'ans'>

$Y_{ij}$ is result for person $j$ belonging to group $i$.

The Model is:  $$Y_{ij}=\mu_i + \epsilon_{ij}$$, where $\epsilon_{ij} \sim N(0, \sigma_i^2)$, $i = 1, 2$ and $j = 1, 2, 3, 4$ and the observations are independent

$i = 1$ is females, otherwise males

$$
\begin{aligned}
H_0: & \sigma_1^2 = \sigma_2^2\\
H_1: & \sigma_1^2 < \sigma_2^2
\end{aligned}
$$

The test statistics is, $$F = \frac{S_\text{males}^2}{S_\text{females}^2} = 3.29$$ 

If we apply 5% level of significance, then we reject $H_0$ if $F > 9.28$. 

Conclusion: We cannot reject.

</div>

b) Based on the result of the test in a) state a model and test if there is difference in statistical knowledge between males/females. The test gave a 0.65 as P-value. Explain in detail what this means.

<div class = 'ans'>

Wa assume equal variances for males and females end state the following model:

$$ Y_{ij} = \mu_i + \epsilon_{ij}$$ where, $\epsilon_{ij} \sim N(0, \sigma^2)$

The hypothesis is,

$$
\begin{aligned}
H_0: & \mu_1 = \mu_2 \\
H_1: & \mu_1 \ne \mu_2
\end{aligned}
$$

```{r}
ttest <- mixlm::t_test(df[1, ], df[2, ], var.equal = T)
```

Assume $H_0$, and $T_\text{calculated} = `r round(ttest$statistic, 3)`$.

Very small $T$, it is impossible to reject $H_0$ also p-value = `r round(ttest$p.value, 3)`.

If males and felmale in average  (population) obtain the same result, the probability of observing a difference (absolute value) between the sample means equal or greater than `r round(ttest$statistic, 3)` is `r round(ttest$p.value, 3)`.

</div>