## Exercise 6: Two sample t-test or One way ANOVA in R
Load the data svin.RData. The data are as described in Week1 slide 17. Final weights have been collected for 9 pigs in the soya group and 8 pigs in the non-soya group. The question addressed is: Do the different feeding strategies give different carcass weight (Norwegian: Slaktevekt)?

```{r}
load("_data/svin.RData")
```


  a) Produce relevant summary statistics and plots.
There are several ways to summarise, e.g., here’s default for _R: Data > Numerical summaries_ split on groups

<div class = 'ans'>

```{r}
sumry <- sapply(list(mean = mean, sd = sd, var = var, n = length), function(x){
  with(soya, {
    tapply(slaktevekt, group, x, simplify = FALSE)
  })
})
knitr::kable(sumry, format = 'html')
```

**Boxplot:** <br />
```{r}
with(soya, plot(slaktevekt ~ group, horizontal = T))
```

</div>

  b) Formulate the model (include assumptions, also equal variances in both groups) and hypotheses.
  
<div class = 'ans'>
  See [Week 1, Exercise - 4 ](#exercise-4) Answer
</div>  
  
  c) Perform the test and comment

<ol type = "I">

<li>As a two-sample test (_R: Statistics > Means ..._)
    
<div class = 'ans'>
  
```{r}
mixlm::t_test(slaktevekt ~ group, data = soya, var.equal = TRUE)
```

</div>
</li>
<li>As an ANOVA

<div class = 'ans'>

```{r}
mixlm::anova_reg(lm(slaktevekt ~ group, data = soya))
```

**Comment:** Same p-value. ANOVA only gives two sided test.

</div>    
</li>
</ol>
  d) Calculate a 95% confidence interval for the difference in weights in the two
  groups.
  
<div class = 'ans'>

See [Week 1, Exercise - 6(c)](#exercise-6)

</div>
  
  e) Repeat c) and d) above but now use a significance level of 0.01 and calculate 99%
  confidence intervals.
  f) Explain the difference between one-sided and two-sided hypotheses.