## Exercise 4

Assume the following result from a completely randomized block design (called experiment1)
Where 

```{r}
y <- matrix(c(1, 3, 2, 5), 2, 2, dimnames = list(Treatment = list(1, 2), Block = list(1, 2)))
mdlExp1 <- lm(value ~ Treatment + Block, data = reshape2::melt(y))
aovExp1 <- unclass(summary(aov(mdlExp1)))[[1]]
```

<div class = "column-3">
|               |       | Block |       |
|---------------|-------|-------|-------|
| **Treatment** |       | **1** | **2** |
|               | **1** | 1     | 2     |
|               | **2** | 3     | 5     |
</div>


- State the model.

Use pocket calculator (if necessary).

- State the model for the block experiment.
- Find SSTotal, SSTReatment, SSBlock and SSError.
- Estimate all the parameters in the model.
- Find the standard error of the estimates.
- Find the residuals.
- Find the fitted values.
- Split the fitted values into a general part, a part by treatment and a part by block.
- Make an ANOVA table and, test if there is treatment effect. 

_You can check your results with R-commander_

If you instead obtained the following result (Experiment 2)

```{r}
y <- matrix(c(1, 4, 2, 6), 2, 2, dimnames = list(Treatment = list(1, 2), Block = list(1, 2)))
mdlExp2 <- lm(value ~ Treatment + Block, data = reshape2::melt(y))
aovExp2 <- unclass(summary(aov(mdlExp2)))[[1]]
```

<div class = "column-3">
|               |       | Block |       |
|---------------|-------|-------|-------|
| **Treatment** |       | **1** | **2** |
|               | **1** | 1     | 2     |
|               | **2** | 4     | 6     |
</div>


Answer the following without any calculations.

Experiment 2 gave:

- larger SSTR than Experiment 1?
- larger SSB than Experiment 1?
- larger SSE than Experiment 1?
- larger SST than Experiment 1?
- larger F or a larger p-value when testing for treatment effect than Experiment 1.

Analyze experiment 1 (incorrectly) by a one factor model.

- Why is SSE larger compared to the block design?
- Is the unexplained variance ($\sigma^2$) larger or small compared to the block design?

### From textbook

#### Exercise 4.2 
The	ANOVA	from	a	randomized	complete	block	experiment	output	is	shown	below.

```{r, echo = FALSE}
anova4.2 <- data.frame(
  Source = c('Treatment', 'Block', 'Error',  'Total'),
  DF = c(4, '?', 20, 29),
  SS = c(1020.56, '?', 169.33, 1513.71),
  MS = c('?', 64.765, '?', ''),
  F = c(30.14, '?', '', ''),
  P = c('?', '?', '', '')
)
knitr::kable(anova4.2)
```

(a) Fill	in	the	blanks.		You	may	give	bounds	on	the	P-value.

    <div class = 'ans'>
    
    ```{r}
    anova4.2 <- data.frame(
      Source = c('Treatment', 'Block', 'Error',  'Total'),
      DF = c(4, 5, 20, 29),
      SS = c(1020.56, 323.82, 169.33, 1513.71),
      MS = c(255.14, 64.765, 8.4665, NA),
      "F" = c(30.14, 7.64956, NA, NA),
      P = c('< 0.0001',  0.00037, '', '')
    )
    gsub("NA", "  ", knitr::kable(anova4.2))
    ```
    
    We can find p-value using computer or some advanced calculator but try to find a range of p-value from the F-table.
    </div>
    
(b) How	many	blocks	were	used	in	this	experiment?
    
    <div class = 'ans'>
    One block factor with 5 levels is used in this experiment.
    </div>


(c) What	conclusions	can	you	draw?


### Problems and solutions for previous exams are found on fronter.

- Exam Stat 210 Sep 5, 2011.
- Exam Stat 210 Sep, 2012 Exercise 2.
- Exam Stat 210 Sep, 2013 Exercises 1 and 2.
- Exam Stat 210 Sep, 2014 Exercise 2.

**Comment:** Datasets (named `Exam*.RData`) are available for the exam problems for those who would like to reproduce output and check answers using R.