# Week Two

```{r, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(comment = NA, echo = FALSE)
```

## Exercise 1

Air pollution measured by chlorine content (measured in ppm, parts per million) for three cities, randomly selected during one year. 

One year ago the cities were approximately equal polluted, but City 1 has the last year tried to reduce the pollution. Data is in a word file called _Monday week 2_ on fronter.

```{r, message=FALSE, warning=FALSE}
txt <- "
C1	C2	C3
0.46	1.21	0.44
0.92	10.86	0.71
0.86	2.09	2.87
0.33	6.76	0.81
0.53	3.20	7.34
5.13	2.02	2.21
1.51	2.16	9.00
1.00	0.44	17.20
0.70	7.80	2.76
1.89	1.20	1.07
0.68	1.39	2.07
0.71	2.42	1.70
0.41	1.62	7.89
0.78	10.78	0.79
"
con <- textConnection(txt)
Pollution <- read.table(con, header = T)
close(con)
library(mixlm)
```


a) Stack the data. Use the ANOVA model to investigate if the chlorine content really differs. Check the model assumptions, which problems do you see?

    <div class = 'ans'>
    
    One of the way you can import data from word file into R is using _clipboard_. Just copy the data and import the data in clipboard into R (Rcommander: _Data > Import Data > from text file ..._)
    
    ```{r}
    PolutStack <- `names<-`(stack(Pollution), c("Chlorine", "City"))
    ```
    
    Let $y_{ij}$ be the chlorine content (polution measurement) for city $i$. The ANOVA model to investigate if the chlorine content really differs is,
    
    \begin{equation}
      y_{ij} = \mu + \tau_i + \epsilon_{ij}, \text{ where, } \epsilon_{ij} \sim \text{NID}\left(0, \sigma^2\right)
      (\#eq:W2E1Model)
    \end{equation}
    
    Since, city is randomly selected, we also have assumption that the  treatment effect follows normal distribution with mean 0 and variance $\sigma_\tau^2$, i.e. $\tau_i\sim\text{NID}\left(0, \sigma_\tau^2\right)$
    
    The ANOVA table for the Model (\@ref(eq:W2E1Model)) is,
    
    ```{r}
    mdl <- mixlm::lm(Chlorine ~ r(City), data = PolutStack)
    CityAnova <- unclass(summary(aov(mdl)))[[1]]
    gsub("NA", "  ", knitr::kable(CityAnova, digits = 3))
    ```
    
    The p-value in ANVOA table is larger than 0.05, so can not reject following Hypothesis at 95% confidence level and stay on the fact that the polution level of different cities are same. However, at 90% confidence level the cities appears to be significantly different.
    
    $$
    \begin{aligned}
    H_0 &: \sigma_\tau ^ 2 = 0 \\
    H_1 &: \sigma_\tau ^ 2 > 0
    \end{aligned}
    $$
    
    Following diagnostic plot (Rcommander: _Model>Graph>Basic Diagnostic Plot_) gives a picture of assumption wheather they hold true in this situation.
    
    ```{r, fig.height = 6}
    op <- par(mfrow = c(2, 2), oma = c(0, 0, 3, 0))
    plot(mdl, c(1, 2, 4, 5), add.smooth = FALSE)
    par(op)
    ```
    
    **Try to interpreate the plot youself**
    
    </div>
    

b) Take the logarithm of the response. (_Data > Manage variables in active dataset > compute new variables_). Note that natural logarithm is done by `log(variable)`. Repeat the analysis, but on `log(chlorine)`. Check model assumptions once again. Investigate if city 1 has been able to reduce the pollution significantly compared the two others cities by a suitable contrast.

    <div class = 'ans'>
    
    ```{r}
    PolutStack$`log(Chlorine)` <- log(PolutStack$Chlorine)
    mdl2 <- mixlm::lm(`log(Chlorine)` ~ r(City), data = PolutStack)
    
    cityAnova2 <- unclass(summary(aov(mdl2)))[[1]]
    gsub("NA", "  ", knitr::kable(cityAnova2, digits = 4))
    ```
    
    The diagnostic plot can be obtained as in previous question,
    
    ```{r, fig.height = 6}
    op <- par(mfrow = c(2, 2), oma = c(0, 0, 3, 0))
    plot(mdl2, c(1, 2, 4, 5), add.smooth = FALSE)
    par(op)
    ```
    
    **Interpreate the plot youself and find its differences from the previous diagnostic plot**
    
    To investigate if `city 1` has been able to reduce the pollution significantly compared the two others cities, a contrast with cofficient `c(1, -0.5, -0.5)`. Since, 
    
    $$\text{Contrast } (\Gamma) = \mu_1 - \frac{1}{2}\left(\mu_2 + \mu_3\right) = \tau_1 - \frac{1}{2}\left(\tau_2 + \tau_3\right)$$ 
    
    The hypothesis for this test is,
    
    $$
    \begin{aligned}
    H_0 &: \Gamma = 0\\
    H_1 &: \Gamma < 0
    \end{aligned}
    $$ 
    
    In R-commander, we can test this hypothesis from _Models > Test Contrasts in ANOVA_. The result for the test is,
    
    ```{r}
    cntrFit <- gmodels::fit.contrast(mdl2, "City", c(1, -0.5, -0.5))
    knitr::kable(cntrFit)
    ```
    
    Since this is one sided hypothesis, the p-value for this test will be _half_ the p-value we obtain in above table ([_See More_](http://www.ats.ucla.edu/stat/mult_pkg/faq/general/tail_tests.htm)). i.e. 
    
    $$\frac{`r round(cntrFit[1, 4], 4)`}{2} = `r round(cntrFit[1, 4] / 2, 4)`$$
    
    Since the p-value is very small, we reject the null hypothesis and claim that `City1` has significantly low chlorine level as compared to the other two cities.
    </div>
    
    
